A formal treatment of bidirectional typing\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}

### 2023-12-21 tex/ESOP24/BiSig.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

\setlength{\marginparwidth}{2cm}
\usepackage[obeyFinal,color=yellow,textsize=scriptsize]{todonotes}

@@

\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
\newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}

@@

%\let\theorem\relax
%\let\definition\relax
%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}

%\author{Liang-Ting Chen
%\orcidID{0000-0002-3250-1331}
%\and 
%Hsiang-Shang Ko
%\orcidID{0000-0002-2439-1048}}
%
%\institute{Institute of Information Science, Academia Sinica, Taiwan}

\title{A formal treatment of bidirectional typing%
%  \thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

>>>>

\setlength{\marginparwidth}{2cm}
\usepackage[color=yellow,textsize=scriptsize]{todonotes}

@@

\SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}
\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
\newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}

@@

%\let\theorem\relax
%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
%\let\definition\relax
%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}

\author{Liang-Ting Chen
\orcidID{0000-0002-3250-1331}
\and 
Hsiang-Shang Ko
\orcidID{0000-0002-2439-1048}}

\institute{Institute of Information Science, Academia Sinica, Taiwan}

\title{A formal treatment of bidirectional typing\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

### 2023-12-21 tex/ESOP24/S1-intro.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ summarised in their survey paper} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should also start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

@@

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
%\Josh{The algorithm adds annotations to the input term?}
can be transformed to a well typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

In more detail:
Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

@@

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account, in particular properly formulating the type synthesis problem.
We also propose a preprocessing step called \emph{mode decoration} that helps the user to work with annotations in practice.

To give a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

>>>>

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
In their survey paper, \varcitet{Dunfield2021}{ summarised} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

@@

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one: traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
%\Josh{The algorithm adds annotations to the input term?}
can be transformed to a well-typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

@@

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.

The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
Therefore, to solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provides information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

### 2023-12-21 tex/ESOP24/S2-outline.tex
[ tex ] Update Future work
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
  \end{mathpar}
  \caption{Typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-typing-derivations}
\end{figure}

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms for simply typed $\lambda$-calculus are defined%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

One (less important) benefit of mode decoration is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

The construction is straightforward:
Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
\begin{theorem}[Soundness and Completeness]
$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
\end{theorem}
%
%
%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
%\end{lemma}

We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their general definitions, which we formulate next.

>>>>

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms for simply typed $\lambda$-calculus are defined%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.

@@

\begin{figure}
  \small
  \bgroup
  \renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ r l }
    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
  \end{tabular}
  \egroup
  \centering
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
  \end{mathpar}
  \caption{Typing derivations for simply typed $\lambda$-calculus}
  \label{fig:STLC-typing-derivations}
\end{figure}
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

One (less important) benefit of mode decoration is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations: observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

The construction is straightforward:
run the mode decorator on the input term~$\erase\Gamma \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
\begin{theorem}[Soundness and Completeness]
$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
\end{theorem}
%
%
%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
%\end{lemma}

We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove the above theorem~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
To quantify over all such systems, we need their general definitions, which we formulate next.
